---
title: "Analysis_Just-Birds"
author: "Marina"
date: "11/16/2021"
output: html_document
---

### install packages and data
```{r packages and data, message=FALSE}
library(tidyverse)
library(mgcv)
library(sf)
library(lwgeom)
library(lme4)
library(ggsci)
library(ggrepel)
library(stats)


setwd("~/GitHub/desert-fires-impact-biodiversity")

animals <- st_read("shapefiles/animals/animals_51022.shp") %>% 
  st_drop_geometry() %>% 
  rename(obsID = TARGET_FID, order = order_, individualCount = individual, obsDay = day, obsMonth = month, obsYear = year, institutionalCode = institutio, treatmentGroup = treatmentG, fireName = INCIDENT, fireYear = FIRE_YEARn, fireSize = GISAcres) %>% 
  select(-c("verbatimSc", "issue", "coordinate", 'nearRd_FID', "nearUA_FID")) %>% 
  filter(obsYear > 1999, obsYear < 2021, species != "Canis lupus", species != "Balaenoptera musculus", species != "Grus canadensis", class == "Aves")
  

ndvi_burned <- read_csv("data/mean_ndvi_burned.csv") %>% 
  mutate(treatmentGroup = "burned")
ndvi_nb <- read_csv("data/mean_ndvi_never_burned.csv") %>% 
  mutate(treatmentGroup = "never burned")

ndvi <- rbind(ndvi_burned, ndvi_nb) %>% 
  rename(obsYear = year)

data <- animals %>%
  group_by(obsYear, treatmentGroup, desert) %>% 
  summarize(totalObs = n()) %>% 
  filter(treatmentGroup != "burned prior 2000")

```

### species occurrences reported

```{r tables and graphs for methods part 2of3 - class composition of data, message=FALSE}

## which species make up the occurrences
species_dist <- animals %>% group_by(species) %>% 
  filter(treatmentGroup != "burned prior 2000") %>% 
  summarize(total_rec = n()) %>% 
  arrange(desc(-total_rec)) %>% 
  mutate(percent = (total_rec/sum(total_rec)),label = (paste0(species, "\n", total_rec)))

unique(species_dist$species)
  
tail(species_dist)
head(species_dist)

```

## Standardize observations by area and add annual NDVI

```{r areas, include=FALSE}
fireid <- unique(animals$fireID)

fire_area <- st_read("shapefiles/fires/fires2000.shp") %>% #polygons for never-burned desert sites
        st_transform(crs = 4326) %>% #transform data to wgs84 to match raster
        filter(fireID %in% fireid)

fire_area_sj <- fire_area %>% 
  filter(desert == "San Joaquin")
fire_area_moj <- fire_area %>% 
  filter(desert == "Mojave")
fire_area_son <- fire_area %>% 
  filter(desert == "Sonoran")


desert <- c("San Joaquin", "Mojave", "Sonoran")
area_km2 <- c(as.numeric(sum(fire_area_sj$area_km2)),
              as.numeric(sum(fire_area_moj$area_km2)),
              as.numeric(sum(fire_area_son$area_km2)))

areas <- data.frame(desert, area_km2) %>% 
  mutate(treatmentGroup = "burned")

head(areas)

nb_area <- st_read("shapefiles/deserts/minus-fires/deserts-minus-fires.shp") %>% #polygons for never-burned desert sites
        st_transform(crs = 4326) #transform data to wgs84 to match raster

nb_area_sj <- nb_area %>% 
  filter(desert == "San Joaquin")
nb_area_moj <- nb_area %>% 
  filter(desert == "Mojave")
nb_area_son <- nb_area %>% 
  filter(desert == "Sonoran")

nb_area_moj$area_km2 <- st_geod_area(nb_area_moj) / 1000000
as.numeric(sum(nb_area_moj$area_km2))
nb_area_sj$area_km2 <- st_geod_area(nb_area_sj) / 1000000
as.numeric(sum(nb_area_sj$area_km2))
nb_area_son$area_km2 <- st_geod_area(nb_area_son) / 1000000
as.numeric(sum(nb_area_son$area_km2))

area_km2 <- c(as.numeric(sum(nb_area_sj$area_km2)),
              as.numeric(sum(nb_area_moj$area_km2)),
              as.numeric(sum(nb_area_son$area_km2)))
areas_nb <- data.frame(desert, area_km2) %>% 
  mutate(treatmentGroup = "never burned")
areas <- rbind(areas, areas_nb)

head(areas)

```

### Animal occurrences standardized to # of occurrences reported per 10,000km^2 

```{r NDVI to data, warnings=FALSE}
data <- animals %>%
  group_by(obsYear, treatmentGroup, desert) %>% 
  summarize(totalObs = n()) %>% 
  filter(treatmentGroup != "burned prior 2000")

#add ndvi data
data <- left_join(data, ndvi, by = c("desert", "obsYear", "treatmentGroup"))

#add area for burned and unburned
data <- left_join(data, areas, by = c("desert", "treatmentGroup")) %>% #areas is a dataframe created from shapefiles in a hidden step
  mutate(stzd_totalObs = (totalObs * 10000)/area_km2) #total observation per 10000 km2

data$desert <- as.factor(data$desert)
data$treatmentGroup <- as.factor(data$treatmentGroup)

data %>% group_by(treatmentGroup, desert) %>% summarize(totalObs = sum(totalObs), mean_ndvi = mean(ndvi), area_km2 = mean(area_km2), total_standard_obs = sum(stzd_totalObs))


ggplot(ndvi, aes(x = obsYear, y = ndvi, color = treatmentGroup, shape = desert)) +
  geom_point() +
  ylim(0, 1) +
  xlim(2000, 2020) +
  facet_grid(vars(treatmentGroup)) +
  ylab("ndvi") +
  scale_color_manual(values = c("firebrick4", "skyblue4"))

  
ggplot(ndvi, aes(x = treatmentGroup, y = ndvi, color = desert)) +
  geom_boxplot(outlier.shape = 18)  +
  ylab("Annual mean NDVI") +
  xlab("") +
  scale_color_uchicago("dark") +
  theme_classic() + 
  theme(axis.text.y = element_text(size = 12), 
        axis.text.x = element_text(size = 12),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12)) +
  scale_x_discrete(labels=c("Burned area", "Control"))


```

### how many total occurrences reported for each desert over the 20-year period?

```{r viz on occurrences over the years}

data %>% group_by(treatmentGroup, desert) %>% summarize(totalObs = sum(totalObs), mean_ndvi = mean(ndvi), area_km2 = mean(area_km2), total_standard_obs = sum(stzd_totalObs)) %>% 
  ggplot(aes(x = treatmentGroup, y = total_standard_obs, fill = desert)) +
  geom_col(position = "dodge") +
  ylab(bquote('Total occurrences reported per 10,000 km'^2)) +
  xlab("")+
  scale_fill_uchicago("dark") +
  theme(axis.text.y = element_text(size = 12), 
        axis.text.x = element_text(size = 12),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12)) +
  scale_x_discrete(labels=c("Burned area", "Control"))
```

### how have the number of occurrences changed throughout the years?

```{r}

ggplot(data, aes(x = obsYear, y = stzd_totalObs, color = treatmentGroup)) +
  geom_point(size = 2) +
  geom_smooth(method = "gam") +
  ylim(0, 600) +
  xlim(2000, 2020) +
  facet_grid(vars(desert)) +
  ylab(bquote('Total occurrences reported per 10,000 km'^2)) +
  xlab("Year") +
  scale_color_manual(values = c("#b3697a", "#697ab3"),
                     name = "Burn Treatment",
                     labels = c("Burned", "Control")) +
  scale_fill_manual(values = c("#b3697a", "#697ab3"),
                    name = "Burn Treatment",
                    labels = c("Burned", "Control")) +
  theme_bw()

```

### linear model
 

### lms and GAMs


```{r}
model_lm1 <- lm(stzd_totalObs ~ obsYear + ndvi + desert + treatmentGroup, data = data)
summary(model_lm1)

model_lm2 <- lm(stzd_totalObs ~ obsYear + ndvi + treatmentGroup, data = data)
summary(model_lm2)

model_lm3 <- lm(stzd_totalObs ~ obsYear + desert + ndvi, data = data)
summary(model_lm3)

AIC(model_lm1, model_lm2, model_lm3)

model_gam0 <- gam(stzd_totalObs ~ s(obsYear) + ndvi + desert + treatmentGroup, data = data, method = "REML")
summary(model_gam0)

model_gam1 <- gam(stzd_totalObs ~ s(obsYear) + s(ndvi) + desert + treatmentGroup, data = data, method = "REML")
summary(model_gam1)

model_gam2 <- gam(stzd_totalObs ~ s(obsYear) + ndvi + desert + treatmentGroup,family = tw(link = "log"), data = data, method = "REML")
summary(model_gam2)

model_gam3 <- gam(stzd_totalObs ~ s(obsYear) + s(ndvi) + desert + treatmentGroup, family = tw(link = "log"), data = data, method = "REML")
summary(model_gam3)

gam.check(model_gam0)
gam.check(model_gam1)
gam.check(model_gam2)
gam.check(model_gam3)


AIC(model_gam0, model_gam1, model_gam2, model_gam3, model_lm1)
```

model_gam3 has a lower AIC value and is, therefore, a better-fit model.

### plot GAMs

```{r}
plot(model_gam3, all.terms = TRUE)

```




### ANOVA test

```{r}
head(data)

occ_anova1 <- aov(stzd_totalObs ~ treatmentGroup, data = data)
occ_anova2 <- aov(stzd_totalObs ~ desert + treatmentGroup, data = data)
occ_anova3 <- aov(stzd_totalObs ~ desert*treatmentGroup, data = data)

AIC(occ_anova1, occ_anova2, occ_anova3)

summary(occ_anova3)

TukeyHSD(occ_anova3)

plot(occ_anova3)

```

Results: 
- significant difference between occurrences reported in San Joaquin and Mojave.
- no sig. diff. between burned and control (neither overall or between deserts)


