---
title: "EDA Endangered Animals"
author: "Marina"
date: "10/7/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)

setwd("~/GitHub/desert-fires-impact-biodiversity")

```

### Load libraries and data

```{r}
library(tidyverse)
library(skimr)

sites <- read_csv("data/sitesv2-2022.csv") %>% 
  filter(fireYear <= 2017)
observations <- read_csv("data/observations-2022.csv") %>% 
  rename(obsYear = year, obsMonth = month) %>% 
  filter(obsYear < 2021)

#remove duplicates based on coordinates and individual count and species
observations <- observations[!duplicated(observations[,c(9:20)]),]

data <- merge(observations, sites[,c("fireID", "fireMonth", "startDate", "endDate")], by = "fireID")

data$fireID <- as.character(data$fireID)

head(data)

#check species
unique(data$species)

#remove Canis familiaris (domestic dog) erroneously include by GBIF when search for Canis lupis
data <- subset(data, species != "Canis familiaris")

```

## Pre-/Post- value

To compare observation before a fire and after a fire, create a pre-/post- column that identifies the observation as occurring either before a fire or after a fire. 

"postingvalue" = obsYear - fireYear
"pre_or_post" = if postingvalue is negative = pre; positive = post; 0 = SY\
those in the same year will be further catergorized by comparing the month of the observation.
"n_years" = absolute value of years between observation and fire



```{r}
data <- data %>% 
  mutate(postingvalue = obsYear - fireYear, 
         pre_or_post = case_when(postingvalue > 0 ~ "post",
                                 postingvalue < 0 ~ "pre",
                                 postingvalue == 0 ~ "SY"),
         n_years = abs(postingvalue))
  
#find pre- post- for observations in the same year and replace SY in datafire with pre or post
SY <- filter(data, postingvalue == 0) %>% 
  mutate(postingmonth = obsMonth - fireMonth,
    pre_or_post = case_when(postingmonth >0 ~ "post",
                            postingmonth < 0 ~ "pre",
                            postingvalue == 0 ~ "SM"))

data$pre_or_post <- ifelse(data$pre_or_post == "SY", SY$pre_or_post, data$pre_or_post)

unique(data$pre_or_post) #check if any SM (same month); would need to further categorize

```
### Need observations both before and after a fire. Need at least 2 observations before and 2 observations after. This code block just does it not taking number of years between observation and fire into account.

```{r, filter sites by observation before and after}
# need atleast 2 observation pre and 2 observation post

pre_and_post <- group_by(data, fireID, pre_or_post) %>%
  summarize(total = n()) %>% 
  spread(pre_or_post, total) %>%
  mutate(post = ifelse(is.na(post), 0, post), pre = ifelse(is.na(pre), 0, pre)) %>%
  filter(post > 1 & pre > 1)

two_each <- group_by(data, fireID, obsYear, pre_or_post) %>% 
  summarise(total = n()) %>% 
  filter(fireID %in% pre_and_post$fireID)

view(two_each)

#filter only fires that have observations before and after a fire
data2 <- data %>%
  filter(fireID %in% pre_and_post$fireID)

```


## At least 1 observation within 4 years before and at least 1 within 4 years after a fire.

```{r}
pre_and_post_4y <- data %>% filter(n_years <= 4) %>% 
  group_by(fireID, pre_or_post) %>%
  summarize(total = n()) %>% 
  spread(pre_or_post, total) %>%
  mutate(post = ifelse(is.na(post), 0, post), pre = ifelse(is.na(pre), 0, pre)) %>%
  filter(post > 0 & pre > 0)

two_each <- group_by(data, fireID, obsYear, pre_or_post) %>% 
  summarise(total = n()) %>% 
  filter(fireID %in% pre_and_post$fireID)

view(two_each)

#filter only fires that have observations before and after a fire
data2 <- data %>%
  filter(fireID %in% pre_and_post$fireID)

```



```{r}
skim(data)
```

Compare pre to post

```{r}
total_observations <- data %>%
  group_by(fireID, pre_or_post) %>% 
  summarize(totalObs = n())

total_observations

data <- left_join(data, total_observations[, c("fireID", "totalObs", "pre_or_post")], by = c("fireID", "pre_or_post"))

ggplot(data, aes(reorder(pre_or_post, totalObs), totalObs)) +
  geom_violin(fill = "grey75") +
  geom_boxplot(alpha = 0.5, width = 0.1) + 
  stat_summary(fun.y=mean, geom="point", shape=18, size=5, color="red", fill="red") +
  xlab("") +
  ylab("Total Occurrences") +
  theme_bw(base_size = 20)

stats <- data %>% group_by(pre_or_post) %>% 
  summarize(mean = mean(totalObs), min = min(totalObs),
            max = max(totalObs))
stats


```

observations over the years

```{r}

posting <- data %>% 
  group_by(fireID, postingvalue, pre_or_post, n_years, desert, class) %>% 
  filter(postingvalue >-6, postingvalue <6) %>% 
  summarize(total = n())

mean(posting$total)
var(posting$total)


model1 <- glm(total ~ postingvalue, family = "quasipoisson", posting)
model2 <- glm(total ~ postingvalue + desert, family = "quasipoisson", posting)
model3 <- glm(total ~ postingvalue + desert + class, family = "quasipoisson", posting)

summary(model1)
summary(model2)
summary(model3)

ggplot(posting, aes(postingvalue, total, color = pre_or_post)) + 
  geom_point() + 
  geom_smooth(method = "glm", method.args = list(family = "quasipoisson")) +
  ylab("Total occurrences") +
  xlab("Posting value") +
  scale_color_manual(values = c("red3", "blue4")) +
  theme_bw(base_size = 20) + theme(legend.title = element_blank())

```

Total observations per site before and after a fire

```{r}
data_by_fire <- data %>% 
  group_by(fireID, pre_or_post) %>% 
  summarize(total = n())


ggplot(data_by_fire, aes(pre_or_post, total)) +
  geom_violin(fill = "olivedrab") +
  geom_boxplot(alpha = 0.5, width = 0.1) + 
  stat_summary(fun.y=mean, geom="point", shape=18, size=5, color="red", fill="red") +
  labs(y = "observations") +
  scale_x_discrete(limits = c("pre", "post"))

```




